Fractional Types

### 2016-12-19 typespp.tex
global revisions; sec 5 is the only remaining thing to do

\begin{definition}[$\Pi$]
\label{def:pi}
The syntax of $\Pi$ is given by the following categories:
\[\begin{array}{lrcl}
(\textrm{Types}) &
  \tau &::=& \zt \alt \ot \alt \tau_1 \oplus \tau_2 \alt \tau_1 \otimes \tau_2 \\
(\textrm{Values}) &
  v &::=& \unitv \alt \inl{v} \alt \inr{v} \alt (v_1,v_2) \\
(\textrm{1-combinators}) &
  c &:& \tau_1 \iso \tau_2 ~ [\textit{see Fig.~\ref{pi-combinators}}] \\
(\textrm{2-combinators}) &
  \alpha &:& c_1 \isotwo c_2 \mbox{~where~} c_1, c_2 : \tau_1 \iso \tau_2
  ~[\textit{see Fig.~\ref{pi-combinators2}}]
\end{array}\]
\end{definition}

\noindent Both classes of programs, 1-combinators $c$, and
2-combinators~$\alpha$, denote \emph{equivalences} in the Homotopy Type
Theory (HoTT) sense~\cite{hottbook}. The
elements $c$ of 1-combinators denote type isomorphisms. The elements $\alpha$ of
2-combinators denote the set of sound and complete equivalences between these
type isomorphisms. Using the 1-combinators, it is possible to write any
reversible boolean function and hence encode arbitrary boolean functions by a
technique that goes back to Toffoli~\cite{Toffoli:1980}. The 2-combinators
provide a layer of programs that computes semantics-preserving transformations
of 1-combinators. As a small example, let us abbreviate $\ot \oplus \ot$ as the
type $\mathbb{2}$ of booleans and examine two possible implementations of
boolean negation. The first directly uses the primitive combinator
$\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to exchange the two
values of type~$\mathbb{2}$; the second use three consecutive uses of $\swapp$
to achieve the same effect:
\[\begin{array}{rcl}
\mathsf{not_1} &=& \swapp \\
\mathsf{not_2} &=& (\swapp \odot \swapp) \odot \swapp
\end{array}\]
We can write a 2-combinator whose \emph{type} is $\mathsf{not_2}
\isotwo \mathsf{not_1}$:
\[
(\linvdl ~\respstwo~ \idisotwo)~\transtwo~\idldl
\]
which not only shows the equivalence of the two implementations of negation but
also shows \emph{how} to transform one to the other. This rewriting focuses
on the first two occurrences of $\swapp$ and uses $\linvdl$ to reduce them to
$\idiso$ since they are inverses. It then uses $\idldl$ to simplify the
composition of $\idiso$ with $\swapp$ to just $\swapp$.

@@

For plain sets, the definition just counts the elements as each element is its
own equivalence class and has exactly one automorphism (the identity). Without
quite formalizing them and relying on the informal diagrams until the next
section, we argue that each of the groupoids $G_1$, $G_2$, and $G_3$ in
Fig.~\ref{fig:groupoids2} has cardinality $\frac{3}{2}$. Groupoid~$G_1$ consists
of two isomorphism classes: class~$a$ has one object with one automorphism (the
identity) and class~$c$ has one object with two distinct automorphisms; the
cardinality is $\frac{1}{1} + \frac{1}{2} = \frac{3}{2}$. For groupoid~$G_2$, we
also have two isomorphism classes with representatives $a$ and $c$; the class
containing $a$ has two automorphisms starting from $a$: the identity and the
loop going from $a$ to~$b$ and back. By the groupoid axioms, this loop is
equivalent to the identity which means that the class containing $a$ has just
one automorphism. The isomorphism class of $c$ has two non-equivalent
automorphisms on it and hence the cardinality of $G_2$ is also
$\frac{1}{1} + \frac{1}{2} = \frac{3}{2}$. For~$G_3$, we have three isomorphism
classes, each with two non-equivalent automorphisms, and hence the cardinality
of $G_3$ is $\frac{1}{2} + \frac{1}{2} + \frac{1}{2} = \frac{3}{2}$.  It is
important to note that $G_1$ and $G_2$ are (categorically) equivalent groupoids,
but that $G_3$ is not equivalent to either $G_1$ or $G_2$.  Roughly speaking
this is because the number of connected components is also a categorical
invariant of a groupoid, and here $G_1$ and $G_2$ have $2$ whilst $G_3$ has $3$.

@@

The previous construction is quite useful: it allows us to take a set of
cardinality $N$ and a permutation on that set of order $P$ to construct a
groupoid of cardinality $\frac{N}{P}$. Although this idea allows us to construct
a groupoid of cardinality $\frac{3}{2}$ as shown above, it is not expressive
enough to construct a groupoid of cardinality, say, $\frac{1}{3}$. Indeed if the
underlying set has only one element, the only permutation is the identity and
$P$ must be 1.

@@

\begin{definition}[$\iorder{p}$] For each 1-combinator $p : \tau\iso\tau$, we form the 2-groupoid $\iorder{p}$ as follows:
\begin{itemize}
\item the objects are the iterates of the identity combinator on $\tau$;
\item the morphisms between every pair of objects are the elements of $\iter{p}$;
\item there is a 2-morphisms between between 1-morphisms $\triple{k_1}{q_1}{\alpha_1}$ and $\triple{k_2}{q_2}{\alpha_2}$ for each $\alpha : q_1 \isotwo q_2$ for every $\alpha : q_1 \isotwo q_2$
\end{itemize}
\end{definition}

@@

The looping groupoid $\looping{G}$ has the elements $g$ of the group $G$ as
objects and morphisms between elements in the same conjugacy class; that is,
there is a
morphism between elements $g$ and $h$ iff there exists a $x \in G$ such that
$g = x^{-1} \cdot g \cdot x$. Fixing a type $\tau$ and a particular 1-combinator
$p : \tau \iso \tau$, we get a group whose elements are the iterates $p^k$. The
corresponding looping groupoid has all the iterates as objects, and morphisms
between $p^{k_1}$ and $p^{k_2}$ whenever there exists an integer $j$ such that
$p^{k_2}$ is equivalent to $p^{-j} \odot p^{k_1} \odot p^{j}$. By
Lem.~\ref{lem:distiterplus}, the latter term is $\isotwo$-equivalent to
$p^{-j+k_1+j}$ which is $\isotwo$-equivalent to $p^{k_1}$. In other words, we
have a morphism between $p^{k_1}$ and $p^{k_2}$ whenever they are
$\isotwo$-equivalent which is consistent with our definition of iteration
groupoids.

@@

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
\[
\order{p} \circledast \iorder{p} \simeq \order{\idiso}
\]
The left hand side is the categorical product of the iteration groupoid and
symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
groupoid is 1. The right hand side is the iteration groupoid for the identity
1-combinator which also has cardinality 1. The groupoids on either side are
however not isomorphic, there do not exist full and faithful functors between
them, and hence $\simeq$ cannot be categorical equivalence.

There are however weaker notions of groupoid equivalence, such as \emph{Morita
  equivalence}~\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that identify
the two groupoids above.  Although the notion of equivalence we use in the next
section appears novel, it is related ``in spirit'' to Morita equivalence and we
find it useful to give a small example. We will verify that the product of the
groupoid $\order{\permfive}$ of cardinality 3 and the groupoid
$\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is Morita equivalent to the
trivial groupoid with only one distinct object and one trivial automorphism. To
avoid clutter, we assume the groupoids are strict, e.g., that the groupoid
$\order{\permfive}$ has three elements (instead of 3 isomorphism classes of
elements). Getting an equivalence in this case reduces to finding a \emph{Morita
  morphism} from $\order{\idiso}$ to
$\order{\permfive} \circledast \iorder{\permfive}$. This is a functor $\psi$
from the unit groupoid to the product groupoid, say $G$, with an additional
condition that the commutative diagram in figure~\ref{fig:morita} is a
\emph{pullback}, where $s,t$ are the \emph{source} and \emph{target} maps, and
$\ot$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids defined
previously.  The intuition is that the codomain consists of three objects with
associated morphisms, each of them counting as a ``third.'' The functor from
$\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
include an arbitrary choice of ``which third'' to target and the universality
condition of the pullback ensures that the choice is ultimately insignificant as
it leads to the same result as every other possible functor.

To summarize, the equivalence between the two groupoids requires a \emph{local}
map that chooses a target and a \emph{global} condition ensuring that the choice
ultimately annihilates. The challenge we are facing is to turn this idea into an
operational semantics that somehow combines the local and global aspects of the
equivalence. We will be inspired by the folklore result in categorical semantics
of dependent type theory, that ``substitution into a dependent type is a
pullback''~\cite{subpullback}, (but which turns out to have been fully explained
in~\cite{practicalfoundations}) as explained in the next section.

\begin{figure}[h]
  \centering
  \begin{tikzcd}
    Q
    \arrow[drr, bend left, "{\xi_1}"]
    \arrow[ddr, bend right, "{(s,t)}"]
    \arrow[dr, dotted] & & \\
    & \order{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
    & G \arrow[d, "{(s,t)}"] \\
    & \ot \times \ot \arrow[r, "{\psi_0 \times \psi_0}"]
    & \mathbb{3} \times \mathbb{3}
  \end{tikzcd}
  \caption{
    \label{fig:morita}
    Pullback diagram for a Morita morphism,
    $\order{\idiso} \rightarrow \order{\permfive} \circledast
    \iorder{\permfive}$
  }
\end{figure}

@@

%%%%%
\subsection{First attempt}

@@

To understand the challenge in designing an operational semantics for $\eta$ and
$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
manipulation on types:
\[
x \mapsto (1 \times x) \mapsto ((x \times 1/x) \times x)
\mapsto (x \times (1/x \times x)) \mapsto (x \times 1) \mapsto x
\]
and by the coherence conditions of compact closed categories, we expect this
circuit, when run in either direction, to be equivalent to the
identity. Consider now the case of a combinator $c$ whose order is greater than
or equal to 2, i.e., the type $\order{c}$ has at least two distinct values
$\idiso$ and $c$ and the type $\iorder{c}$ has at least two distinct values
$1/\idiso$ and $1/c$. We require this circuit to produce $\idiso$ when given
$\idiso$ from either end, and similarly to produce $c$ when given $c$ from
either end. The problem becomes apparent when one observes that the uses of
$\eta+$ in the left-to-right execution and the use of $\epsilon-$ in the
right-to-left execution must behave differently depending on the incoming
value. But the ``local'' information available to either of these combinators
does not include the ``global'' information about the incoming value and there
is no way for them to make consistent guesses locally. It is possible to modify
the operational semantics to use a mechanism to express a computational effect
which enables spatially separated parts of the program to communicate in a way
that is reminiscent of entanglement in quantum mechanics. Possibilities to
realize this communication are global reference cells, backtracking, logical
variables with unification. We present in the next section a more abstract
approach that encodes the required dependency in dataflow constraints using
dependent types.

\begin{figure}[t]
\begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
  \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
  \node at (0.5,1) {\footnotesize{$\unitetl/$}};
  \path (-1.1,1) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (0,1);
  \path (1,1.75) edge node[above] {\footnotesize{$\order{\idiso}$}} (1.6,1.75);
  \path [highlight] (1,0.25) edge node[above] {} (4,0.25);
  \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (4,0.25);
  \draw (1.6,1) -- (2.6,1) -- (2.6,2.8) -- (1.6,2.8) -- cycle;
  \node at (2.1,1.9) {\footnotesize{$\eta+$}};
  \draw[dashed] (2.8,2.6) -- (2.8,1.25);
  \path [highlight] (2.6,2.6) edge node[above] {} (4,2.6);
  \path (2.6,2.6) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (4,2.6);
  \path [highlight] (2.6,1.25) edge node[above] {} (4,1.25);
  \path (2.6,1.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (4,1.25);
  \draw (4,0) -- (5.5,0) -- (5.5,3) -- (4,3) -- cycle;
  \node at (4.75,1.5) {$\footnotesize{\assocrt/}$};
  \path [highlight] (5.5,1.25) edge node[above] {} (7,1.25);
  \path (5.5,1.25) edge node[above] {$\footnotesize{1/\#c}$} (7,1.25);
  \path [highlight] (5.5,2.6) edge node[above] {} (9,2.6);
  \path (5.5,2.6) edge node[above] {\footnotesize{$\#c$}} (9,2.6);
  \draw (7,0) -- (8,0) -- (8,1.6) -- (7,1.6) -- cycle;
  \node at (7.5,0.8) {\footnotesize{$\epsilon-$}};
  \draw[dashed] (6.8,0.2) -- (6.8,1.25);
  \path (8,1) edge node[above] {\footnotesize{$\order{\idiso}$}} (9,1);
  \path [highlight] (5.5,0.2) edge node[above] {} (7,0.2);
  \path (5.5,0.2) edge node[above] {\footnotesize{$\#c$}} (7,0.2);
  \draw (9,0.8) -- (10,0.8) -- (10,2.8) -- (9,2.8) -- cycle;
  \node at (9.5,1.8) {\footnotesize{$\unitetr/$}};
  \path (10,1.8) edge node [above] {\footnotesize{$\#c$}} (11,1.8);
\end{tikzpicture}
\caption{\label{fig:zigzag} Zig-zag}
\end{figure}

@@

The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
where each type $\mathbb{T}$ refers to a particular value $v$ in
$\mathbb{T}$. As all fractional types are inhabited (see Remark
in Sec.~\ref{sec:symmetryg}), there is no loss in generality in going
to pointed types. We then allow universal quantification over indices $k$
associated to values, over a collection of types.  In particular, the type
$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
imposes a constraint on the two values associated with the types $\order{c}$ and
$\iorder{c}$; they must be ``synchronized'' by sharing the same~$k$ but are
otherwise arbitrary. Once we introduce a universal quantifier in one direction
of execution, we are led to also include an existential quantifier for the
reverse execution. Intuitively, a forward evaluation step that is at liberty to
choose a $k$ will, when reversed encounter a particular, i.e., existentially
quantified, $k$. The syntax is therefore as follows.

@@

A type $\mathbb{T}_\bullet$ is generally of the form
$\forall\exists{\overline{k}}.[\mathbb{T},v]$ where all the quantifiers are
lifted to the top leaving a non-dependent type and a value in that type. The
combinators in Fig.~\ref{pifracd:comb} also serve as the operational semantics
for the language as they explain how the value currently in focus changes during
evaluation. Given the new combinators, we can modify the the circuit in
Fig.~\ref{fig:zigzag} by adding explicit synchronization primitives. The
evaluation proceeds as follows where we disambiguate the uses of $\forall$ and
$\exists$ for clarity.

Forward execution:
\[\begin{array}{rcl}
[\order{c},c^i]
&\stackrel{\unitetl/}{\leftrightblackspoon}&
  [\order{id}\circledast\order{c}, \langle \idiso,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\leftrightblackspoon}&
  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
&\stackrel{\synchl}{\leftrightblackspoon}&
  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
&\stackrel{\assocrt/}{\leftrightblackspoon}&
  [\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
&\stackrel{\packr}{\leftrightblackspoon}&
  \exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\leftrightblackspoon}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\unitetr/}{\leftrightblackspoon}&
  [\order{c}, c^i]
\end{array}\]

>>>>

\begin{definition}[$\Pi$]
\label{def:pi}
The syntax of $\Pi$ is given by the following categories:
\[\begin{array}{lrcl}
(\textrm{Types}) &
  \tau &::=& \zt \alt \ot \alt \tau_1 \oplus \tau_2 \alt \tau_1 \otimes \tau_2 \\
(\textrm{Values}) &
  v &::=& \unitv \alt \inl{v} \alt \inr{v} \alt (v_1,v_2) \\
(\textrm{1-combinators}) &
  c,p &:& \tau_1 \iso \tau_2 ~ [\textit{see Fig.~\ref{pi-combinators}}] \\
(\textrm{2-combinators}) &
  \alpha &:& c_1 \isotwo c_2 \mbox{~where~} c_1, c_2 : \tau_1 \iso \tau_2
  ~[\textit{see Fig.~\ref{pi-combinators2}}]
\end{array}\]
\end{definition}

\noindent Both classes of programs, 1-combinators $c$, and
2-combinators~$\alpha$, denote \emph{equivalences} in the Homotopy Type Theory
(HoTT) sense~\cite{hottbook}. The elements $c$ or $p$ of 1-combinators denote
type isomorphisms. The elements $\alpha$ of 2-combinators denote the set of
sound and complete equivalences between these type isomorphisms. Using the
1-combinators, it is possible to write any reversible boolean function and hence
encode arbitrary boolean functions by a technique that goes back to
Toffoli~\cite{Toffoli:1980}. The 2-combinators provide a layer of programs that
computes semantics-preserving transformations of 1-combinators. As a small
example, let us abbreviate $\ot \oplus \ot$ as the type $\mathbb{2}$ of booleans
and examine two possible implementations of boolean negation. The first directly
uses the primitive combinator
$\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to exchange the two
values of type~$\mathbb{2}$; the second uses three consecutive $\swapp$s to
achieve the same effect:
\[\begin{array}{rcl}
\mathsf{not_1} &=& \swapp \\
\mathsf{not_2} &=& (\swapp \odot \swapp) \odot \swapp
\end{array}\]
We can write a 2-combinator whose \emph{type} is $\mathsf{not_2}
\isotwo \mathsf{not_1}$:
\[
(\linvdl ~\respstwo~ \idisotwo)~\transtwo~\idldl
\]
which not only shows the equivalence of the two implementations of negation but
also shows \emph{how} to transform one to the other. This rewriting focuses
on the first two occurrences of $\swapp$ and uses $\linvdl$ to reduce them to
$\idiso$ since they are inverses. It then uses $\idldl$ to simplify the
composition of $\idiso$ with $\swapp$ to just $\swapp$.

@@

For plain sets, the definition just counts the elements as each element is its
own equivalence class and has exactly one automorphism (the identity). Without
quite formalizing them and relying on the informal diagrams until the next
section, we argue that each of the groupoids $G_1$, $G_2$, and $G_3$ in
Fig.~\ref{fig:groupoids2} has cardinality $\frac{3}{2}$. Groupoid~$G_1$ consists
of two isomorphism classes: class~$a$ has one object with one automorphism (the
identity) and class~$c$ has one object with two distinct automorphisms; the
cardinality is $\frac{1}{1} + \frac{1}{2} = \frac{3}{2}$. For groupoid~$G_2$, we
also have two isomorphism classes with representatives $a$ and $c$; the class
containing $a$ has two automorphisms starting from $a$: the identity and the
loop going from $a$ to~$b$ and back. By the groupoid axioms, this loop is
equivalent to the identity which means that the class containing $a$ has just
one automorphism. The isomorphism class of $c$ has two non-equivalent
automorphisms on it and hence the cardinality of $G_2$ is also
$\frac{1}{1} + \frac{1}{2} = \frac{3}{2}$. For~$G_3$, we have three isomorphism
classes, each with two non-equivalent automorphisms, and hence the cardinality
of $G_3$ is $\frac{1}{2} + \frac{1}{2} + \frac{1}{2} = \frac{3}{2}$.  It is
important to note that $G_1$ and $G_2$ are categorically equivalent groupoids,
but that $G_3$ is not categorically equivalent to either $G_1$ or $G_2$.  Roughly speaking
this is because the number of connected components is also a categorical
invariant of a groupoid, and here $G_1$ and $G_2$ have $2$ whilst $G_3$ has $3$.

@@

The previous construction known in the literature as an \emph{action
  groupoid}~\cite{groupoidintro} is quite useful: it allows us to take a set of
cardinality $N$ and a permutation on that set of order $P$ to construct a
groupoid of cardinality $\frac{N}{P}$. Although this idea allows us to construct
a groupoid of cardinality $\frac{3}{2}$ as shown above, it is not expressive
enough to construct a groupoid of cardinality, say, $\frac{1}{3}$. Indeed, if the
underlying set has only one element ($N=1$) the only permutation is the identity
and $P$ must be 1.

@@

\begin{definition}[$\iorder{p}$] For each 1-combinator $p : \tau\iso\tau$, we form the 2-groupoid $\iorder{p}$ as follows:
\begin{itemize}
\item the objects are the iterates of the identity combinator on $\tau$;
\item the morphisms between every pair of objects are the elements of $\iter{p}$;
\item there is a 2-morphism between 1-morphisms $\triple{k_1}{q_1}{\alpha_1}$ and $\triple{k_2}{q_2}{\alpha_2}$ for every $\alpha : q_1 \isotwo q_2$.
\end{itemize}
\end{definition}

@@

The looping groupoid $\looping{G}$ has the elements $g$ of the group $G$ as
objects and morphisms between elements in the same conjugacy class; that is,
there is a
morphism between elements~$g$ and $h$ iff there exists an $x \in G$ such that
$h = x^{-1} \cdot g \cdot x$. Fixing a type $\tau$ and a particular 1-combinator
$p : \tau \iso \tau$, we get a group whose elements are the iterates $p^k$. The
corresponding looping groupoid has all the iterates as objects, and morphisms
between $p^{k_1}$ and $p^{k_2}$ whenever there exists an integer $j$ such that
$p^{k_2}$ is equivalent to $p^{-j} \odot p^{k_1} \odot p^{j}$. By
Lem.~\ref{lem:distiterplus}, the latter term is $\isotwo$-equivalent to
$p^{-j+k_1+j}$ which is $\isotwo$-equivalent to $p^{k_1}$. In other words, we
have a morphism between $p^{k_1}$ and $p^{k_2}$ whenever they are
$\isotwo$-equivalent which is consistent with our definition of iteration
groupoids.

@@

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The integration of these groupoids with the categorical sum is
complicated and left for future work. For products, the most prominent identity,
from which many other identities follow, is:
\[
\order{p} \circledast \iorder{p} \simeq \order{\idiso}
\]
The left hand side is the categorical product of the iteration groupoid and
symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
groupoid is 1. The right hand side is the iteration groupoid for the identity
1-combinator which also has cardinality 1. The groupoids on either side are
however not isomorphic, there do not exist full and faithful functors between
them, and hence $\simeq$ cannot be categorical equivalence.

There are however weaker notions of groupoid equivalence, such as \emph{Morita
  equivalence}~\cite{looping}\cite[C5.3]{johnstone2002sketches}, that identify
the two groupoids above.  Although the notion of equivalence we use in the next
section appears novel, it is related ``in spirit'' to Morita equivalence and we
find it useful to give a small example. We will verify that the product of the
groupoid $\order{\permfive}$ of cardinality 3 and the groupoid
$\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is Morita equivalent to the
trivial groupoid with only one distinct object and one trivial automorphism. To
avoid clutter, we assume the groupoids are strict, e.g., that the groupoid
$\order{\permfive}$ has three elements (instead of 3 isomorphism classes of
elements).

Getting an equivalence in this case reduces to finding a \emph{Morita morphism}
from $\order{\idiso}$ to $\order{\permfive} \circledast
\iorder{\permfive}$.
This is a functor $\psi$ from the unit groupoid to the product groupoid, say
$G$, with an additional condition that the commutative diagram below is a
\emph{pullback}, where $s,t$ are the \emph{source} and \emph{target} maps, and
$\ot$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids defined
previously.

\begin{center}
\begin{tikzcd}
    Q
    \arrow[drr, bend left, "{\xi_1}"]
    \arrow[ddr, bend right, "{(s,t)}"]
    \arrow[dr, dotted] & & \\
    & \order{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
    & G \arrow[d, "{(s,t)}"] \\
    & \ot \times \ot \arrow[r, "{\psi_0 \times \psi_0}"]
    & \mathbb{3} \times \mathbb{3}
\end{tikzcd}
\end{center}

\medskip\noindent The intuition is that the codomain consists of three objects
with associated morphisms, each of them counting as a ``third.'' The functor
from $\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
include an arbitrary choice of ``which third'' to target and the universality
condition of the pullback ensures that the choice is ultimately insignificant as
it leads to the same result as every other possible functor.

To summarize, the equivalence between the two groupoids requires a \emph{local}
map that chooses a target and a \emph{global} condition ensuring that the choice
ultimately annihilates. The challenge we aim to solve in the next section is to
turn this idea into an operational semantics that somehow combines the local and
global aspects of the equivalence. We will be inspired by the folklore result in
the categorical semantics of dependent type theory~\cite{subpullback}, first
explained by Paul Taylor~\cite{practicalfoundations}, that ``substitution into a
dependent type is a pullback''.

@@

%%%%%
\subsection{First attempt: non-dependent version}

@@

To understand the challenge in designing an operational semantics for $\eta$ and
$\epsilon$ in the non-dependent setting, we consider the ``zigzag'' circuit in
Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
manipulation on types:
\[
x \mapsto (1 \times x) \mapsto ((x \times 1/x) \times x)
\mapsto (x \times (1/x \times x)) \mapsto (x \times 1) \mapsto x
\]
and by the coherence conditions of compact closed categories, we expect this
circuit, when run in either direction, to be equivalent to the
identity. Consider now the case of a combinator $c$ whose order is greater than
or equal to 2, i.e., the type $\order{c}$ has at least two distinct values
$\idiso$ and $c$ and the type $\iorder{c}$ has at least two distinct values
$1/\idiso$ and $1/c$. We require this circuit to produce $\idiso$ when given
$\idiso$ from either end, and similarly to produce $c$ when given $c$ from
either end. The problem becomes apparent when one observes that the uses of
$\eta+$ in the left-to-right execution and the use of $\epsilon-$ in the
right-to-left execution must behave differently depending on the incoming
value. But the ``local'' information available to either of these combinators
does not include the ``global'' information about the incoming value and there
is no way for them to make consistent guesses locally. It is possible to design
an operational semantics that uses computational effects to enable spatially
separated parts of the program to communicate in a way that is reminiscent of
entanglement in quantum mechanics. Possibilities to realize this communication
are global reference cells, backtracking, and logical variables with
unification. We choose however to present in the next section a more abstract
approach that encodes the required dependency in dataflow constraints using
dependent types.

\begin{figure}[t]
\begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
  \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
  \node at (0.5,1) {\footnotesize{$\unitetl/$}};
  \path (-1.1,1) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (0,1);
  \path (1,1.75) edge node[above] {\footnotesize{$\order{\idiso}$}} (1.6,1.75);
  \path [highlight] (1,0.25) edge node[above] {} (4,0.25);
  \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (4,0.25);
  \draw (1.6,1) -- (2.6,1) -- (2.6,2.8) -- (1.6,2.8) -- cycle;
  \node at (2.1,1.9) {\footnotesize{$\eta+$}};
  \draw[dashed] (2.8,2.6) -- (2.8,1.25);
  \path [highlight] (2.6,2.6) edge node[above] {} (4,2.6);
  \path (2.6,2.6) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (4,2.6);
  \path [highlight] (2.6,1.25) edge node[above] {} (4,1.25);
  \path (2.6,1.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (4,1.25);
  \draw (4,0) -- (5.5,0) -- (5.5,3) -- (4,3) -- cycle;
  \node at (4.75,1.5) {$\footnotesize{\assocrt/}$};
  \path [highlight] (5.5,1.25) edge node[above] {} (7,1.25);
  \path (5.5,1.25) edge node[above] {$\footnotesize{1/\#c}$} (7,1.25);
  \path [highlight] (5.5,2.6) edge node[above] {} (9,2.6);
  \path (5.5,2.6) edge node[above] {\footnotesize{$\#c$}} (9,2.6);
  \draw (7,0) -- (8,0) -- (8,1.6) -- (7,1.6) -- cycle;
  \node at (7.5,0.8) {\footnotesize{$\epsilon-$}};
  \draw[dashed] (6.8,0.2) -- (6.8,1.25);
  \path (8,1) edge node[above] {\footnotesize{$\order{\idiso}$}} (9,1);
  \path [highlight] (5.5,0.2) edge node[above] {} (7,0.2);
  \path (5.5,0.2) edge node[above] {\footnotesize{$\#c$}} (7,0.2);
  \draw (9,0.8) -- (10,0.8) -- (10,2.8) -- (9,2.8) -- cycle;
  \node at (9.5,1.8) {\footnotesize{$\unitetr/$}};
  \path (10,1.8) edge node [above] {\footnotesize{$\#c$}} (11,1.8);
\end{tikzpicture}
\caption{\label{fig:zigzag} Zigzag}
\end{figure}

@@

We modify the syntax of $\Pi^/$ as follows.

@@

In the new definition, we enrich the types to pointed types $\mathbb{T_\bullet}$
where each type $\mathbb{T}$ refers to a particular value $v$ in
$\mathbb{T}$. As all fractional types are inhabited (see Remark in
Sec.~\ref{sec:symmetryg}), there is no loss in generality in going to pointed
types. We then allow universal quantification over indices $k$ associated to
values, over a collection of types.  In particular, the type
$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
imposes a constraint on the two values associated with the types $\order{c}$ and
$\iorder{c}$; they must be ``synchronized'' by sharing the same~$k$ but are
otherwise arbitrary. Once we introduce a universal quantifier in one direction
of execution, we are led to also include an existential quantifier for the
reverse execution. Intuitively, a forward evaluation step that is at liberty to
choose a $k$ will, when reversed encounter a particular, i.e., existentially
quantified, $k$. A type $\mathbb{T}_\bullet$ is therefore generally of the form
$\forall\exists{\overline{k}}.[\mathbb{T},v]$ where all the quantifiers are
lifted to the top followed by a non-dependent type and a value in that type.

The adaptation of the combinators to the pointed dependent setting in Fig .~\ref
{pifracd:comb} is relatively straightforward. First, the types of the
combinators must keep track of how the value currently in focus changes during
evaluation, and hence directly encode the operational semantics of the
language. As suggested above, the quantifiers are interpreted differently
depending on which direction the rule is used. We use a $\forall\exists$
notation to express the presence of \emph{one} quantifier that is interpreted
ddifferently in each direction. The notation has the advantage of being compact,
avoiding duplication, and making the symmetry of the rules explicit. We also
include two new pairs of combinators $\synchl/packl$ and $\synchr/packr$ that
serve as explicit quantifier introduction and elimination rules: they serve as
explicit ``synchronization'' points as illustrated in the example below.

The example revisits the zigzag circuit in Fig.~\ref{fig:zigzag} using the
dependent types and the new explicit synchronization primitives. The evaluation
proceeds as follows where we disambiguate the uses of $\forall$ and $\exists$
in each direction for clarity:

\paragraph*{Forward execution of zigzag circuit with synchronization.}
\[\begin{array}{rcl}
[\order{c},c^i]
&\stackrel{\unitetl/}{\leftrightblackspoon}&
  [\order{id}\circledast\order{c}, \langle \idiso,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\leftrightblackspoon}&
  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
&\stackrel{\synchl}{\leftrightblackspoon}&
  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
&\stackrel{\assocrt/}{\leftrightblackspoon}&
  [\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
&\stackrel{\packr}{\leftrightblackspoon}&
  \exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\leftrightblackspoon}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\unitetr/}{\leftrightblackspoon}&
  [\order{c}, c^i]
\end{array}\]
